---
title: "R and Ebola: Meeting 3"
author: "Brooke Anderson"
date: "November 21, 2014"
output: ioslides_presentation
---

## Ebola data

Pull in your ebola data again so we'll have it ready for examples:

```{r}
## If necessary, use setwd() to get to the right directory
ebola <- read.table("country_timeseries.csv", sep = ",",
                    header = TRUE)
ebola[1:3, 1:5]
```

## Two functions to help with loops

Two functions that are very useful for loops are `print()` and `paste()`:

```{r}
print(1:3)
paste("Brooke", "Anderson")
paste("Brooke", "Anderson", sep = ".")
```

## Two functions to help with loops
```{r}
rankings.name <- c("First", "Second", "Third")
rankings.num <- c(1:3)
paste(rankings.name, "is what we call #", rankings.num)
```

## Loop structure

```{r, eval = FALSE}
for([index] in [vector]){     ## Generic code
        [code you want to run for a single loop]
}
```

## Simple loops

```{r}
for(i in 1:3){
        print("For this loop, i is:")
        print(i)
}
```

## Simple loops

```{r}
for(i in 1:3){
        print(paste("For this loop, i is", i))
}
```

## Using `i` to index

Remember how we can use indexing to just get certain parts of a vector or dataframe:

```{r}
my.family <- c("Reeves", "Brooke", "Cord")
my.family[1]
```

## Using `i` to index

We can take advantage of this in loops:

```{r}
for(i in 1:3){
        print(my.family[i])
}
```

This turns out to be very powerful... 

## Using `i` to index

To figure out a loop, try the code inside the loop a few times with `i` equal to different values it would get in the loop:

```{r}
i <- 1
print(my.family[i])
i <- 2
print(my.family[i])
```

## Practice

Now you try:

- Create a vector, `Country` of all the countries in the ebola dataset. Use the `paste()` function to create a vector called `case.colnames` the gives the names of the columns for all countries.
- Create vectors of first and last names of everyone in your group, then write a loop to print "[Name] is in our group" for everyone. 
- Create a loop that prints, on each run, "Die roll # [i] is: [random number between 1 and 6]".

*Hints: 
Create separate `first.name` and `last.name` vectors and use `paste()` in your loop to print each person's full name. 
Use the `sample()` function to get the random number (e.g., `sample(1:6, size = 1)`).*

## Using loops

According to programmers, you should never repeat yourself.

You can use loops to do the same thing a lot of times. For example, we could use a loop to create a new dataframe with, for each country, the mean number of ebola cases.

Basically, we'll create a empty dataframe and use the loop to fill it in.

```{r, echo = FALSE}
Country <- c("Guinea", "Liberia", "SierraLeone", "Nigeria",
             "Senegal", "UnitedStates", "Spain", "Mali")
ebola.cases <- data.frame(country = Country,
                          nonmissing.obs = NA)
ebola.cases
```

## Using loops

First, we'll need a vector of all the column names for cases.

```{r}
Country <- c("Guinea", "Liberia", "SierraLeone", "Nigeria",
             "Senegal", "UnitedStates", "Spain", "Mali")
case.colnames <- paste("Cases", Country, sep = "_")
case.colnames
```

## Using loops

Next, we'll create a dataframe that we'll fill in with the loop:

```{r}
ebola.cases <- data.frame(country = Country,
                          nonmissing.obs = NA)
ebola.cases
```

## Using loops

Now, let's think about what we want to do for each country:

```{r}
mean(ebola[ , "Cases_Guinea"], na.rm = TRUE)
```
Equivalently:
```{r}
i <- 1
case.colnames[i]
mean.cases <- mean(ebola[ , case.colnames[i]], na.rm = TRUE)
mean.cases
```

## Using loops

Once you've calculated what you want, you can use indexing to put the number in your `ebola.cases` dataframe:

```{r}
head(ebola.cases, 2)
ebola.cases[i, "nonmissing.obs"] <- 1
head(ebola.cases, 2)
```

## Using loops

Once you've calculated what you want, you can use indexing to put the number in your `ebola.cases` dataframe:

```{r}
ebola.cases[i, "nonmissing.obs"] <- mean.cases
head(ebola.cases, 2)
```

## Using loops 

Now you can put this all together:

```{r}
Country <- c("Guinea", "Liberia", "SierraLeone", "Nigeria",
             "Senegal", "UnitedStates", "Spain", "Mali")
case.colnames <- paste("Cases", Country, sep = "_")
ebola.cases <- data.frame(country = Country,
                          nonmissing.obs = NA)

for(i in 1:length(Country)){
        mean.cases <- mean(ebola[ , case.colnames[i]], na.rm = TRUE)
        ebola.cases[i, "nonmissing.obs"] <- mean.cases
}
```

## Using loops 

```{r}
head(ebola.cases)
```

## Practice

Now you try:

- Use a loop get the range of each country's number of ebola cases from our data. Put these values in a new dataframe with the columns `Country`, `Min.Cases`, and `Max.Cases`.
- Use a loop to plot ebola cases by day for each of the countries.

## Practice

Step-by-step for first exercise:

- Create a vector called `Country` that lists all of our countries
- Create a vector called `case.colnames` the gives the names of the columns for all countries 
- Create a dataframe called `Case.Ranges` that uses the `Country` vector as a `country` column and then also has columns for `Min.Cases` and `Max.Cases`. To start, these will just be full of `NA`s. 
- Create a loop where `i` goes from 1 to the length of your `Country` vector. 
- For each loop, use `range(ebola, na.rm = TRUE)` to get the range of cases for that country.
- For each loop, use indexing to put this range (it will be two values) in the right places in your `Case.Ranges` dataframe.